---
title: "Introduction to the BayesLinearKinematics package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BayesLinearKinematics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Chunk Options:
# include = FALSE: Prevents code and results from appearing in the finished file.
#                  Useful for setup code.
# collapse = TRUE: Merges R code and its output into a single block,
#                  which is often cleaner for vignettes.
# comment = "#>":  Standard prefix for R output lines. Makes output look like comments.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6.5)

# Load the package
library(BayesLinearKinematics)

# Store package version
pkg_version <- packageVersion("BayesLinearKinematics")

# Store the R version
r_version <- R.version.string
```

This vignette assumes a basic understanding of statistical concepts like expectation, variance, and covariance.

## Introduction

The `BayesLinearKinematics` package provides tools for performing Bayes linear analysis. This involves representing uncertain beliefs about quantities using expectation and covariance, and updating these beliefs based on observed data or other uncertain information (kinematics). This vignette demonstrates the core workflow: defining prior beliefs, structuring observed data, performing adjustments, and evaluating the result.

Note that this vignette was created using R version `r r_version` and the `BayesLinearKinematics` package version `r pkg_version`.

## 1. Defining Prior Beliefs with `bl`

The foundation of a Bayes linear analysis is specifying your prior beliefs about the variables of interest. This is done using the `bl` class, which stores expectations and covariances.

Let's define prior beliefs for three variables, $x$, $y$ and $z$. We expect them to be 1, 2 and 3 respectively. We also specify their covariance structure.

```{r create_bl}
# Define prior beliefs using the bl() constructor
bl_prior <- bl(name = 'Prior Beliefs',        # A descriptive name for this belief set
               varnames = c('x', 'y', 'z'),   # Names of the variables
               expectation = c(1, 2, 3),      # Prior expected values E[x], E[y], E[z]
               covariance = matrix(c(1.0, 0.5, 0.2, # Covariance matrix:
                                     0.5, 1.0, 0.5, #   [Var(x) Cov(y,x) Cov(z,x)]
                                     0.2, 0.5, 1.0),  # [Cov(x,y) Var(y) Cov(z,y)]
                                   nrow = 3, ncol = 3) #[Cov(x,z) Cov(y,z) Var(z)]
              )

# Print the object - uses the custom print method defined for the 'bl' class
# The 'digits' argument controls the rounding in the output
print(bl_prior, digits = 2)
```

Essentially, here, we are encoding:
$$
\begin{align*}
\text{E}\pmatrix{x\\y\\z} &= \pmatrix{1 \\ 2 \\ 3} \\
\text{Cov}\pmatrix{x\\y\\z} &= \pmatrix{1.0 & 0.5 & 0.2 \\ 0.5 & 1.0 & 0.5 \\ 0.2 & 0.5 & 1.0}
\end{align*}
$$

The `print` method gives a clean overview. We can also visualise the covariance matrix using the `plot` method defined for the `bl` class:

```{r plot_bl_prior, fig.show='hold', fig.cap="Plot of the prior covariance matrix."}
# Plot the covariance matrix using the S4 plot method for 'bl' objects
# This method should handle the visualization details internally.
plot(bl_prior)
```

This plot shows the strength and sign of the prior covariances between variables. Red indicates positive covariance, blue negative (though none here), and intensity relates to magnitude. The diagonal elements represent the variances.

## 2. Representing Observed Data with `bl_data`

Suppose we observe values for some of the variables. We represent this using the `bl_data` class. Let's say we observed $x$ to be 1.1 and $y$ to be 2.1. For ease of exposition, we will denote these two observed variables as $D$ (for data).

```{r create_bl_data}
# Define observed data using the bl_data() constructor
observed_data <- bl_data(name = 'Observations',      # A descriptive name for the data
                         varnames = c('x', 'y'),     # Names of the observed variables
                         values = c(1.1, 2.1))       # The observed values

# Print the data object - uses the custom print method for 'bl_data'
print(observed_data)
```

## 3. Adjusting Beliefs using Data (`bl_adjust`)

The core operation is updating our prior beliefs (`bl_prior`) using the `observed_data`. The `bl_adjust` function performs this Bayes linear adjustment according to the standard formulae:
$$
\begin{align*}
\text{E}_D\pmatrix{x\\y\\z} &= \text{E}\pmatrix{x\\y\\z} + \text{Cov}\left[\pmatrix{x\\y\\z} ,\pmatrix{x\\y}\right]\cdot \text{Cov}\pmatrix{x\\y}^{-1} \cdot \left[D - \text{E}\pmatrix{x\\y}\right]\\
\text{Cov}_D\pmatrix{x\\y\\z} &= \text{Cov}\pmatrix{x\\y\\z} - \text{Cov}\left[\pmatrix{x\\y\\z} ,\pmatrix{x\\y}\right]\cdot \text{Cov}\pmatrix{x\\y}^{-1} \cdot \text{Cov}\left[\pmatrix{x\\y\\z} ,\pmatrix{x\\y}\right]^T.
\end{align*}
$$

```{r adjust_beliefs_data}
# Perform the adjustment using the prior beliefs and the observed data
# x: The 'bl' object containing prior beliefs to be adjusted.
# y: The 'bl_data' object containing the observations to adjust by.
bl_adjusted_data <- bl_adjust(x = bl_prior, y = observed_data)

# Print the adjusted beliefs. Note the name is automatically generated.
# Using more digits to see the effect of the adjustment.
print(bl_adjusted_data, digits = 3)
```

Notice how the expectations for $x$, $y$ and $z$ have changed from their prior values (1, 2, 3) based on the observations. The covariance matrix has also been updated, reflecting reduced uncertainty (variances on the diagonal are generally smaller). Let's plot the adjusted covariance:

```{r plot_bl_adjusted_data, fig.show='hold', fig.cap="Plot of the covariance matrix after adjustment with data."}
# Plot the adjusted covariance matrix
plot(bl_adjusted_data)
```
Comparing this plot to the prior covariance plot, we can see changes, particularly a reduction in the variance (diagonal elements) and potentially changes in the off-diagonal covariances, reflecting the information gained from the data.

## 4. Adjusting Beliefs using Kinematics (`bl_adjust`)

The `bl_adjust` function can also update one set of beliefs (`bl` object, `x`) based on another set of related beliefs (`bl` object, `y`). This is sometimes referred to as Bayes linear kinematics. It uses similar formulae but incorporates the uncertainty from the second belief object (`y`).

Let's define a second belief structure, perhaps representing information about $y$ and $z$ from another source (e.g., a different sensor or expert opinion).

```{r create_bl_kinematics}
# Define a second set of beliefs (e.g., from another expert or model)
bl_info <- bl(name = "Sensor Info",
              varnames = c('y', 'z'),          # Variables covered by this info
              expectation = c(2.2, 2.8),      # Expectations from this source
              covariance = matrix(c(0.5, 0.1, # Covariance structure from this source
                                    0.1, 0.6), nrow=2, ncol=2))

print(bl_info, digits=2)
```

Now, we adjust our original `bl_prior` using this `bl_info`. The adjustment combines the information from both sources.

```{r adjust_beliefs_kinematics}
# Perform kinematic adjustment
# x: The 'bl' object containing prior beliefs to be adjusted.
# y: The 'bl' object containing the other beliefs to adjust by.
bl_adjusted_kinematics <- bl_adjust(x = bl_prior, y = bl_info)

# Print the result of the kinematic adjustment
print(bl_adjusted_kinematics, digits=3)

# Plot the resulting covariance after kinematic adjustment
plot(bl_adjusted_kinematics)
```

The adjustment incorporates the information from `bl_info` into the prior beliefs, updating both expectations and covariances according to Bayes linear rules for combining uncertain information.

## 5. Calculating Resolution with `bl_resolution`

How much has our uncertainty been reduced by an adjustment? The `bl_resolution` function calculates this, comparing the prior variance to the adjusted variance for each variable: `Resolution = 1 - Var_adjusted / Var_prior`. This applies when comparing a prior `bl` object (`x`) to an adjusted `bl` object (`y`).

Let's calculate the resolution achieved by the data adjustment performed earlier:

```{r calculate_resolution_data}
# Calculate the resolution comparing the prior to the data-adjusted beliefs
# x: The 'bl' object representing prior beliefs.
# y: The 'bl' object representing adjusted beliefs.
resolution_values_data <- bl_resolution(x = bl_prior, y = bl_adjusted_data)

# Print the resolutions, rounded for clarity
print(round(resolution_values_data, 3))
```

A resolution close to 1 means the variance for that variable was greatly reduced by the adjustment, while a value close to 0 means the adjustment had little impact on its variance. Here, $x$ and $y$ (the observed variables) have higher resolution, indicating their variance decreased significantly. $z$ also has some resolution, reflecting the information gained about it indirectly through its covariance with $x$ and $y$.

We can also calculate the resolution resulting from the kinematic adjustment:
```{r calculate_resolution_kinematics}
# Calculate the resolution comparing the prior to the kinematically-adjusted beliefs
resolution_values_kinematics <- bl_resolution(x = bl_prior, y = bl_adjusted_kinematics)

# Print the resolutions
print(round(resolution_values_kinematics, 3))
```
This shows the variance reduction achieved by combining the `bl_prior` with `bl_info`.

## 6. Subsetting Beliefs with `bl_subset`

Sometimes you might want to focus on a subset of variables from a `bl` or `bl_data` object. The `bl_subset` function allows this, extracting the relevant expectations, values, and sub-matrices of the covariance.

```{r subsetting}
# Extract beliefs only for 'x' and 'z' from the data-adjusted object
# x: The 'bl' or 'bl_data' object to subset.
# varnames: A character vector of variable names to keep.
bl_subset_xz <- bl_subset(bl_adjusted_data, varnames = c('x', 'z'))

print(bl_subset_xz, digits = 3)

# Extract data only for 'x' from the original observations
# (This is a simple case but demonstrates subsetting bl_data)
data_subset_x <- bl_subset(observed_data, varnames = 'x')

print(data_subset_x)
```

## Summary

This vignette demonstrated the basic use of the `BayesLinearKinematics` package:

1.  Create prior beliefs using `bl()`.
2.  Represent observations using `bl_data()`.
3.  Update beliefs using observations or other beliefs via `bl_adjust()`.
4.  Evaluate the information gained (variance reduction) using `bl_resolution()`.
5.  Extract subsets of variables using `bl_subset()`.

These tools provide a foundation for applying Bayes linear methods in R. Remember to consult the help files (e.g., `?bl`, `?bl_adjust`) for further details on each function and its arguments.
