---
title: "Hellinger distances between belief specifications"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Hellinger Distances Between Belief Specifications}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Chunk Options:
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>",
 fig.width = 7,
 fig.height = 6.5)

# Load the package
library(BayesLinearKinematics)

# Load MASS for ginv used in hellinger_squared for multivariate case
# Ensure MASS is listed as a dependency if you haven't already
library(MASS) 

# Store package version
pkg_version <- packageVersion("BayesLinearKinematics")

# Store the R version
r_version <- R.version.string
```

This vignette assumes familiarity with the `bl` object class introduced in the "Introduction to the `BayesLinearKinematics` package" vignette.


Note that this vignette was created using R version `r r_version` and the `BayesLinearKinematics` package version `r pkg_version`.

## Introduction

In Bayes linear analysis, we represent beliefs using expectations and covariances. Sometimes, we want to quantify how "different" two belief specifications are. The Hellinger distance is a metric used to measure the similarity between two probability distributions. While a `bl` object only stores the first two moments (expectation and covariance), we can calculate the Hellinger distance between the multivariate normal distributions that share these moments. This provides a useful measure of discrepancy between two belief specifications encapsulated in `bl` objects.

The squared Hellinger distance $H^2(P, Q)$ ranges between 0 and 1 where 0 indicates the distributions (and thus the expectations and covariances) are identical and 1 indicates maximal difference.

This vignette demonstrates how to use the `hellinger_squared()` function in the `BayesLinearKinematics` package to calculate this distance.

## Defining the Squared Hellinger Distance (Normal Case)

The `hellinger_squared()` function calculates the distance between two `bl` objects `x` and `y` by assuming they represent multivariate normal distributions $P$ and $Q$ respectively with means $\mu_1, \mu_2$ (from `x@expectation`, `y@expectation`) and covariance matrices $\Sigma_1, \Sigma_2$ (from `x@covariance`, `y@covariance`).

Univariate Case: If the `bl` objects represent beliefs about a single variable ($P = N(\mu_1, \sigma_1^2)$, $Q = N(\mu_2, \sigma_2^2)$), the squared Hellinger distance is:
$$H^2(P, Q) = 1 - \sqrt{\frac{2 \sigma_1 \sigma_2}{\sigma_1^2 + \sigma_2^2}} \exp\left(-\frac{1}{4} \frac{(\mu_1 - \mu_2)^2}{\sigma_1^2 + \sigma_2^2}\right)$$

Multivariate Case: If the `bl` objects represent beliefs about multiple variables ($P = N(\mu_1, \Sigma_1)$, $Q = N(\mu_2, \Sigma_2)$), the squared Hellinger distance is:
$$H^2(P, Q) = 1 - \frac{\det(\Sigma_1)^{1/4} \det(\Sigma_2)^{1/4}}{\det\left(\frac{\Sigma_1 + \Sigma_2}{2}\right)^{1/2}} \exp\left(-\frac{1}{8} (\mu_1 - \mu_2)^T \left(\frac{\Sigma_1 + \Sigma_2}{2}\right)^{-1} (\mu_1 - \mu_2)\right)$$
where $\det(\cdot)$ denotes the determinant of a matrix and $(\cdot)^{-1}$ denotes the matrix inverse (or generalised inverse ginv as used in the function for numerical stability).  

## Calculating Distance Between Two Belief Specifications

Let's define two different belief specifications for the same set of variables $x$, $y$ and $z$. We can then use hellinger_squared() to quantify how much these beliefs differ.

First, we define belief_A perhaps representing an initial assessment.

```{r}
# Define first belief specification
belief_A <- bl(name = 'Belief A',
              varnames = c('x', 'y', 'z'),
              expectation = c(1, 2, 3),
              covariance = matrix(c(1.0, 0.5, 0.2,
                                    0.5, 1.0, 0.5,
                                    0.2, 0.5, 1.0),
                                  nrow = 3, ncol = 3))

belief_A
plot(belief_A)
```

Now, we define `belief_B` representing an alternative assessment perhaps from a different expert or model.

```{r}
# Define second belief specification (different expectation and covariance)
belief_B <- bl(name = 'Belief B',
              varnames = c('x', 'y', 'z'),
              expectation = c(1.2, 2.1, 2.9), # Slightly different means
              covariance = matrix(c(1.1, 0.4, 0.3, # Slightly different covariance
                                    0.4, 0.9, 0.6,
                                    0.3, 0.6, 1.2),
                                  nrow = 3, ncol = 3))

belief_B
plot(belief_B)
```

Now, we calculate the squared Hellinger distance between `belief_A` and `belief_B`.

```{r}
# Calculate the squared Hellinger distance
# x: The first 'bl' object.
# y: The second 'bl' object.
dist_A_B <- hellinger_squared(x = belief_A, y = belief_B)

# Print the result
cat(paste("Squared Hellinger distance between Belief A and Belief B:",
          round(dist_A_B, 4)))
```

The resulting value (likely small but non-zero) quantifies the difference between these two belief specifications under the normality assumption. A value closer to 0 indicates higher similarity.

Let's also quickly check the univariate case.

```{r}
# Univariate beliefs
belief_X1 <- bl(name="X1", varnames="x", expectation=1, covariance=matrix(1))
belief_X2 <- bl(name="X2", varnames="x", expectation=1.5, covariance=matrix(1.2))

dist_X1_X2 <- hellinger_squared(belief_X1, belief_X2)

cat(paste("Squared Hellinger distance between Belief X1 and Belief X2:",
          round(dist_X1_X2, 4)))

# Check distance to self (should be 0)
dist_X1_X1 <- hellinger_squared(belief_X1, belief_X1)
cat(paste("\nSquared Hellinger distance between Belief X1 and itself:",
          round(dist_X1_X1, 4)))
```

## Measuring the Impact of Kinematic Adjustment

The Hellinger distance can also be used to measure how much a belief specification changes after undergoing a Bayes linear adjustment (kinematics). We can calculate the distance between the prior beliefs and the adjusted beliefs.

Let's reuse the prior beliefs (`bl_prior`) and the kinematic information (`bl_info`) from the introductory vignette examples.

```{r}
# Define prior beliefs (same as intro vignette)
bl_prior <- bl(name = 'Prior Beliefs',
               varnames = c('x', 'y', 'z'),
               expectation = c(1, 2, 3),
               covariance = matrix(c(1.0, 0.5, 0.2,
                                     0.5, 1.0, 0.5,
                                     0.2, 0.5, 1.0),
                                   nrow = 3, ncol = 3))

# Define kinematic information (same as intro vignette)
bl_info <- bl(name = "Info",
              varnames = c('y', 'z'),
              expectation = c(2.2, 2.8),
              covariance = matrix(c(0.5, 0.1,
                                    0.1, 0.6), nrow=2, ncol=2))

# Perform the kinematic adjustment
bl_adjusted_kinematics <- bl_adjust(x = bl_prior, y = bl_info)

# Quick look at the prior and adjusted beliefs
bl_prior
bl_adjusted_kinematics
```

Now, we calculate the squared Hellinger distance between the prior belief (`bl_prior`) and the belief state after adjustment (`bl_adjusted_kinematics`). Note that `bl_adjust` might only adjust a subset of variables if `bl_info` covers fewer variables than `bl_prior`. `hellinger_squared` requires both inputs to cover the same set of variables. We should calculate the distance over the variables common to both stages after adjustment i.e. the variables present in `bl_adjusted_kinematics`. In this specific case, `bl_adjust` preserves all variables from `bl_prior` so we can compare them directly.

```{r}
# Plot prior covariance
plot(bl_prior)

# Plot adjusted covariance
plot(bl_adjusted_kinematics)

# Calculate squared Hellinger distance between prior and adjusted beliefs
dist_kinematic <- hellinger_squared(x = bl_prior, y = bl_adjusted_kinematics)

# Print the result
cat(paste("\nSquared Hellinger distance between Prior and Adjusted (Kinematic):",
          round(dist_kinematic, 4)))
```

This distance quantifies the magnitude of the belief change induced by the kinematic update from `bl_info`. A larger value implies the kinematic information caused a greater shift in the belief state (mean and/or covariance structure) under the normality assumption.

## Function Notes

The `hellinger_squared` function includes several checks:

*    It ensures both inputs (x and y) are `bl` objects.
*    It verifies that both `bl` objects contain the exact same set of variable names (`varnames`).
*    It internally reorders the second object (y) to match the variable order of the first object (`x`) using `bl_subset` before performing calculations ensuring alignment.

If these conditions are not met, the function will stop and return an informative error message.

## Summary

This vignette demonstrated the use of the hellinger_squared() function to compute a measure of distance between two belief specifications (bl objects) based on their corresponding multivariate normal distributions.

*    We reviewed the formulae used for univariate and multivariate cases.
*    We calculated the distance between two distinct `bl` objects (`belief_A` and `belief_B`).
*    We calculated the distance between a prior belief (`bl_prior`) and its state after a kinematic update (`bl_adjusted_kinematics`) to quantify the impact of the adjustment.

The squared Hellinger distance provides a value between 0 and 1 offering a standardised way to compare belief structures or measure the effect of belief updates within the `BayesLinearKinematics` framework assuming normality based on the first two moments.

